The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor-Couette flow

### 2016-10-04 paper/thingap_mri_paper.tex
captions, edits, Taylor Couette -> Taylor-Couette

\title{The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor Couette flow}
\author{Clark, S.E.\altaffilmark{1}}
\author{Oishi, J.S. \altaffilmark{2,}\altaffilmark{3}}

@@

For matter to accrete from a disk onto a central object, angular momentum must be transported radially outward in the disk. The transport mechanism is likely turbulent, as molecular viscosity alone cannot account for the needed angular momentum transfer, and likely magnetic, as this turbulence is excited even in hydrodynamically stable disks (\citei{Shakura:1973wg}).  Discovered by \citet{Chandrasekhar:1960wh} and \citet{Velikhov:1959} in global geometry, the MRI was subsequent rediscoverd and applied to accretion disks by \citet{Balbus:1991vs}. Since then, the magnetorotational instability (MRI) remains the leading explanation for rapid angular momentum transport in astrophysical disks. The instability in its simplest geometry arises when a differentially rotating disk is threaded by a vertical magnetic field. The presence of the magnetic field linearly destabilizes the disk gas, driving turbulence and angular momentum transport. The MRI likely plays a role in a diverse host of astrophysical systems, including protoplanetary disks and black hole accretion disks, as well as stellar interiors. Despite its importance, many aspects of the MRI remain poorly understood. In particular, the nonlinear saturation mechanism for the MRI is an open question, and a formidable challenge. MRI saturation has been tackled almost exclusively with simulation, with a few notable exceptions detailed below. In this work we investigate the weakly nonlinear behavior of the MRI.

Weakly nonlinear analysis is a perturbative method used to examine the asymptotic behavior of a system near threshold -- that is, when the system is just barely unstable to its most unstable mode. The analytical technique follows the multiscale evolution of fluid variables in a perturbation expansion, allowing the controlled interaction of modes between orders in a perturbation series (see, e.g., Bender \& Orszag for a full mathematical formalism). Weakly nonlinear analysis can be a powerful technique for analytically examining systems which in their full generality exhibit such complicated nonlinear behavior that their study is relegated primarily to the simulation domain. The MRI is one such phenomenon: while there is a rich literature analytically examining the linear MRI, analytical treatments of the \textit{nonlinear} system are relatively few. The weakly nonlinear treatment of the MRI was pioneered by \citeauthor{Umurhan:2007dz} (\citeyear{Umurhan:2007dz}, \citeyear{Umurhan:2007hs}; hereafter URM07). The authors undertook the first weakly nonlinear analysis of the MRI in a thin-gap Taylor Couette flow, and found that the marginal MRI system approaches saturation in a manner analogous to that of Rayleigh-B\'enard convection. Weakly nonlinear analysis was instrumental in our understanding of Rayleigh-B\'enard convection saturation (\citei{Newell:1969}), and the similarities between convection and the thin-gap MRI are the result of important shared symmetries between the systems. The success of URM07 in modeling the MRI system near threshold merits further consideration, but we are unaware of any other attempts to expand upon their theoretical framework. In this work we rigorously rederive the theory of URM07, and expand upon their findings. Our focus here is on fully characterizing the narrow gap MRI system, both for independent theoretical interest and to have a robust comparison point for extensions of this theory into more complicated geometries. In a companion paper we derive for the first time the weakly nonlinear theory of the standard and helical MRI in a wide gap Taylor Couette flow \citep[][hereafter Paper II]{Clark:2016}. The thin- and wide-gap treatments complement one another theoretically, and both are important regimes for comparison with simulation.

This work examines Taylor Couette flow in the thin-gap regime, an idealization in which the radial extent of the channel is very small compared to its distance from the center of rotation, i.e. $(r_2 - r_1) \ll \frac{1}{2} (r_1 + r_2)$ where $r_1$ and $r_2$ are the radii of the inner and outer flow boundaries, respectively. The thin-gap approximation eliminates curvature terms, so the domain geometry is Cartesian rather than cylindrical. The excluded curvature terms have an explicit dependence on $r$, so they make the problem more challenging both analytically and numerically. In particular, in the wide-gap geometry (i.e. true Taylor Couette flow) the base angular velocity is a function of $r$, where in the thin-gap approximation the shear flow reduces to a linear profile.

@@

\begin{figure}[h!]
\centering
\includegraphics[trim=5cm 0cm 0cm 0cm, scale=.35]{setup_diagram.eps}
\caption{Schematic diagram of our set-up, an axisymmetric thin-gap Taylor Couette flow. We investigate a 2D slice of the X-Z (radial-vertical) plane. Our domain is represented by the bolded black box, of width L. The radial dimension is solved with a basis of Chebyshev polynomials, and the vertical dimension is solved on a Fourier basis. }\label{fig:setup}
\end{figure}

@@

%See Figure \ref{figscalings}. 
Note that these are the same scalings as apply to Rayleigh-B\'enard convection and hydrodynamic Taylor Couette flow. Our $x$ dimension, the direction of angular momentum transport, is analogous to the direction of temperature transport in the convection problem. In analogy to these problems, we posit slow variation in both $Z$ and $T$. Each operator is expanded to reflect these scalings -- for instance, $\partial_z$ becomes $\partial_z + \epsilon\partial_Z$. The fast and slow variables are treated as independent in the perturbation series analysis that follows.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/thingap_saturation_mechanism_plots_velocity_slice60.png}
\caption{Top panel: Mean radial profile for $u_y^{saturated} - u_y^0$, the difference between the saturated azimuthal velocity profile and the initial azimuthal velocity profile (TC flow). Bottom panel: each term in the azimuthal velocity steady state force balance (Equation \ref{eqset2} with $\partial_t u_{y} = 0$).} \label{fig:sat_comp_vel}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/thingap_saturation_mechanism_plots_Bfield_slice60.png}
\caption{.}\label{fig:sat_comp_Bfield}
\end{figure}

@@

Figure~\ref{fig:sat_comp_Bfield} shows $B_z$ and the terms corresponding to steady state inductive balance (Equation~\ref{eqset2} with $\partial_t B_{y} = 0$). Here, the instability acts to push magnetic field toward the boundaries in both the bulk and the boundary layers. Note also that the total magnetic energy in the $B_z$ component is significantly larger than the initial condition. \citet{Ebrahimi:2009ey} considered the saturation of a single, strongly super-critical MRI mode allowed to interact non-linearly only with itself and the mean. They considered two important cases, one in which the mean flow was forced to remain at its initial, quasi-Keplerian state for all time, and one in which the background flow was allowed to evolve. This is a crucial difference between the shearing box and our narrow-gap TC flow: perturbations in our simulation can adjust the background flow, whereas in a shearing box, the shear periodicity forbids perturbations from affecting the mean flow. In the case with a freely evolving background flow, \citet{Ebrahimi:2009ey} found a saturated state quite similar to ours: field pushed to the boundaries, and a reduction in shear in the bulk of the flow. Their flows have less pronounced boundary layers, likely because of their much larger $\Pm = 0.1 -1$.

>>>>

\title{The Weakly Nonlinear Magnetorotational Instability in a thin-gap Taylor-Couette flow}
\author{Clark, S.E.\altaffilmark{1}}
\author{Oishi, J.S. \altaffilmark{2,}\altaffilmark{3}}

@@

For matter to accrete from a disk onto a central object, angular momentum must be transported radially outward in the disk. The transport mechanism is likely turbulent, as molecular viscosity alone cannot account for the needed angular momentum transfer, and likely magnetic, as this turbulence is excited even in hydrodynamically stable disks (\citei{Shakura:1973wg}).  Discovered by \citet{Chandrasekhar:1960wh} and \citet{Velikhov:1959} in global geometry, the MRI was subsequently rediscovered and applied to accretion disks by \citet{Balbus:1991vs}. Since then, the magnetorotational instability (MRI) remains the leading explanation for rapid angular momentum transport in astrophysical disks. The instability in its simplest geometry arises when a differentially rotating disk is threaded by a vertical magnetic field. The presence of the magnetic field linearly destabilizes the disk gas, driving turbulence and angular momentum transport. The MRI likely plays a role in a diverse host of astrophysical systems, including protoplanetary disks and black hole accretion disks, as well as stellar interiors. Despite its importance, many aspects of the MRI remain poorly understood. In particular, the nonlinear saturation mechanism for the MRI is an open question, and a formidable challenge. MRI saturation has been tackled almost exclusively with simulation, with a few notable exceptions detailed below. In this work we investigate the weakly nonlinear behavior of the MRI.

Weakly nonlinear analysis is a perturbative method used to examine the asymptotic behavior of a system near threshold -- that is, when the system is just barely unstable to its most unstable mode. The analytical technique follows the multiscale evolution of fluid variables in a perturbation expansion, allowing the controlled interaction of modes between orders in a perturbation series (see, e.g., Bender \& Orszag for a full mathematical formalism). Weakly nonlinear analysis can be a powerful technique for analytically examining systems which in their full generality exhibit such complicated nonlinear behavior that their study is relegated primarily to the simulation domain. The MRI is one such phenomenon: while there is a rich literature analytically examining the linear MRI, analytical treatments of the \textit{nonlinear} system are relatively few. The weakly nonlinear treatment of the MRI was pioneered by \citeauthor{Umurhan:2007dz} (\citeyear{Umurhan:2007dz}, \citeyear{Umurhan:2007hs}; hereafter URM07). The authors undertook the first weakly nonlinear analysis of the MRI in a thin-gap Taylor-Couette flow, and found that the marginal MRI system approaches saturation in a manner analogous to that of Rayleigh-B\'enard convection. Weakly nonlinear analysis was instrumental in our understanding of Rayleigh-B\'enard convection saturation (\citei{Newell:1969}), and the similarities between convection and the thin-gap MRI are the result of important shared symmetries between the systems. The success of URM07 in modeling the MRI system near threshold merits further consideration, but we are unaware of any other attempts to expand upon their theoretical framework. In this work we rigorously rederive the theory of URM07, and expand upon their findings. Our focus here is on fully characterizing the narrow gap MRI system, both for independent theoretical interest and to have a robust comparison point for extensions of this theory into more complicated geometries. In a companion paper we derive for the first time the weakly nonlinear theory of the standard and helical MRI in a wide gap Taylor-Couette flow \citep[][hereafter Paper II]{Clark:2016}. The thin- and wide-gap treatments complement one another theoretically, and both are important regimes for comparison with simulation.

This work examines Taylor-Couette flow in the thin-gap regime, an idealization in which the radial extent of the channel is very small compared to its distance from the center of rotation, i.e. $(r_2 - r_1) \ll \frac{1}{2} (r_1 + r_2)$ where $r_1$ and $r_2$ are the radii of the inner and outer flow boundaries, respectively. The thin-gap approximation eliminates curvature terms, so the domain geometry is Cartesian rather than cylindrical. The excluded curvature terms have an explicit dependence on $r$, so they make the problem more challenging both analytically and numerically. In particular, in the wide-gap geometry (i.e. true Taylor-Couette flow) the base angular velocity is a function of $r$, where in the thin-gap approximation the shear flow reduces to a linear profile.

@@

\begin{figure}[h!]
\centering
\includegraphics[trim=5cm 0cm 0cm 0cm, scale=.35]{setup_diagram.eps}
\caption{Schematic diagram of our set-up, an axisymmetric thin-gap Taylor-Couette flow. We investigate a 2D slice of the X-Z (radial-vertical) plane. Our domain is represented by the bolded black box, of width L. The radial dimension is solved with a basis of Chebyshev polynomials, and the vertical dimension is solved on a Fourier basis. }\label{fig:setup}
\end{figure}

@@

%See Figure \ref{figscalings}. 
Note that these are the same scalings as apply to Rayleigh-B\'enard convection and hydrodynamic Taylor-Couette flow. Our $x$ dimension, the direction of angular momentum transport, is analogous to the direction of temperature transport in the convection problem. In analogy to these problems, we posit slow variation in both $Z$ and $T$. Each operator is expanded to reflect these scalings -- for instance, $\partial_z$ becomes $\partial_z + \epsilon\partial_Z$. The fast and slow variables are treated as independent in the perturbation series analysis that follows.

@@

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/thingap_saturation_mechanism_plots_velocity_slice60.png}
\caption{Top panel: Mean radial profile for $u_y^{saturated} - u_y^0$, the difference between the saturated azimuthal velocity profile and the initial azimuthal velocity profile (TC flow). Bottom panel: each term in the steady state force balance (Equation \ref{eqset2} with $\partial_t u_{y} = 0$). Saturated quantities are computed with $\alpha_{saturation} = \sqrt{b/c}$ and $\epsilon = 0.5$. The saturated state shows reduced shear in the bulk of the flow, outside of the boundary layers.} \label{fig:sat_comp_vel}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{../figures/thingap_saturation_mechanism_plots_Bfield_slice60.png}
\caption{Top panel: Mean radial profile for $B_z^{saturated}$, the saturated vertical magnetic field (black line). $B_z^0 = 1$ is the constant background magnetic field (gray line). Bottom panel: each term in the steady state inductive balance ($\partial_x$ of Equation \ref{eqset4} with $\partial_t B_y = 0$). Saturated quantities are computed with $\alpha_{saturation} = \sqrt{b/c}$ and $\epsilon = 0.5$. The saturated field is pushed to the radial domain boundaries.}\label{fig:sat_comp_Bfield}
\end{figure}

@@

Figure~\ref{fig:sat_comp_Bfield} shows $B_z$ and the terms corresponding to steady state inductive balance ($\partial_x$ of Equation~\ref{eqset4} with $\partial_t B_{y} = 0$). Here, the instability acts to push magnetic field toward the boundaries in both the bulk and the boundary layers. Note also that the total magnetic energy in the $B_z$ component is significantly larger than the initial condition. \citet{Ebrahimi:2009ey} considered the saturation of a single, strongly super-critical MRI mode allowed to interact non-linearly only with itself and the mean. They considered two important cases, one in which the mean flow was forced to remain at its initial, quasi-Keplerian state for all time, and one in which the background flow was allowed to evolve. This is a crucial difference between the shearing box and our narrow-gap TC flow: perturbations in our simulation can adjust the background flow, whereas in a shearing box, the shear periodicity forbids perturbations from affecting the mean flow. In the case with a freely evolving background flow, \citet{Ebrahimi:2009ey} found a saturated state quite similar to ours: field pushed to the boundaries, and a reduction in shear in the bulk of the flow. Their flows have less pronounced boundary layers, likely because of their much larger $\Pm = 0.1 -1$.
